---
title: "Tidy Tuesday - Wealth Inequality"
author: "Katie Kotasek"
date: "2/10/2021"
output: html_document
---
This week, for Tidy Tuesday, I wanted to focus on one type of charting: Distribution charts. Therefore, I created different distribution charts for the following data sets:

1. *Violin* chart for the data set - student_debt
2. *Density* chart for the data set - retirement
3. *Histogram* chart for the data set - home_owner
4. *Boxplot* chart for the data set - race_wealth
5. *Ridgeline* chart for the data set - income_mean

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# required packages
library(ggridges)
library(ggdark)
library(tidyverse)
library(mmbtools)
library(plyr)

mmb_load_fonts()

options(scipen = 999)
```

```{r}
# data sets
# Average family student loan debt for aged 25-55, by race and year normalized to 2016 dollars
student_debt <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')
# Average family liquid retirement savings normalized to 2016 dollars.
retirement <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/retirement.csv')
# Home ownership percentage for families
home_owner <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/home_owner.csv')
# Family wealth by race/year/measure normalized to 2016, with measures of central tendency with mean and median.
race_wealth <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/race_wealth.csv')
# Family-level income by percentile and year.
income_time <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_time.csv')
# Mean income received by each fifth and top 5% of each racial group.
income_mean <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_mean.csv')
```

```{r student loan debt}
# CLEAN UP STUDENT_DEBT DATA
student_debt <- student_debt %>%
  mutate(loan_pct = round((loan_debt_pct * 100), 2)) %>%
  select(year, race, loan_pct)

# VIOLIN PLOT
student_debt %>%
  ggplot(aes(x = race, y = loan_pct, fill = race)) +
    geom_violin() +
    scale_fill_manual(values = c("#003685", "#78BE21", "#008EAA")) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of student loan debt percentage by race",
         subtitle = "Years 1989 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "",
         y = "Loan Debt (%)")
```

```{r retirement}
mu <- ddply(retirement, "race", summarise, grp.mean=mean(retirement))
# DENSITY PLOT
retirement %>%
  ggplot(aes(x = retirement, fill = race)) +
    geom_density() +
    geom_vline(data = mu, aes(xintercept=grp.mean), linetype="dashed") +
    facet_grid(race ~ .) +
    scale_fill_manual(values = c("#003685", "#78BE21", "#008EAA")) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of average retirement savings by race",
         subtitle = "Years 1989 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "Amount in retirement savings ($)",
         y = "Density")
```

```{r home owner}
# CLEAN UP HOME_OWNER DATA
home_owner <- home_owner %>%
  mutate(home_pct = round((home_owner_pct * 100), 2)) %>%
  select(year, race, home_pct)

# HISTOGRAM PLOT
home_owner %>%
  ggplot(aes(x = home_pct, fill = race)) +
    geom_histogram(alpha = 0.9, color = "black") +
    facet_grid(race ~ .) +
    scale_fill_manual(values = c("#003685", "#78BE21", "#008EAA")) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of home owners percentage by race",
         subtitle = "Years 1976 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "Percent home owner (%)",
         y = "Count")
```

```{r race wealth}
# CLEAN UP RACE_WEALTH DATA
rw_avg <- race_wealth %>%
  filter(type == "Average") %>%
  #mutate_at(vars(wealth_family), replace_na, 0) %>%
  select(year, race, wealth_family)
rw_med <- race_wealth %>%
  filter(type == "Median") %>%
  #mutate_at(vars(wealth_family), replace_na, 0) %>%
  select(year, race, wealth_family)

# BOX PLOT - AVERAGE
rw_avg %>%
  ggplot(aes(x = race, y = wealth_family, fill = race)) +
    geom_boxplot(alpha = 0.8) +
    scale_fill_manual(values = c("#003685", "#78BE21", "#FFC845", "#008EAA")) +
    #geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of average family wealth by race",
         subtitle = "Years 1963 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "",
         y = "Average wealth ($)")
# BOX PLOT - MEDIAN
rw_med %>%
  ggplot(aes(x = race, y = wealth_family, fill = race)) +
    geom_boxplot(alpha = 0.8) +
    scale_fill_manual(values = c("#003685", "#78BE21", "#FFC845", "#008EAA")) +
    #geom_jitter(color="black", size=0.4) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of median family wealth by race",
         subtitle = "Years 1963 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "",
         y = "Median wealth ($)")
```

```{r income time, eval=FALSE, echo=FALSE}
income_time %>%
  ggplot(aes(x = income_family, y = percentile, fill = income_family)) +
    geom_density_ridges() +
    #scale_fill_viridis(name = "income_family", option = "C") +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Income by percentile",
         subtitle = "Years 1963 - 2016",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "Income ($)",
         y = "Percentile")
```

```{r income mean}
# CLEAN UP INCOME_MEAN DATA
income_mean <- income_mean %>%
  mutate(race = case_when(race == "All Races" ~ "All",
                          race == "Asian Alone or in Combination" ~ "Asian",
                          race == "Asian Alone" ~ "Asian",
                          race == "Black Alone or in Combination" ~ "Black",
                          race == "Black Alone" ~ "Black",
                          race == "Hispanic" ~ "Hispanic",
                          race == "White Alone" ~ "White",
                          race == "White, Not Hispanic" ~ "White (NH)")
         )
income_curr <- income_mean %>%
  filter(dollar_type == "Current Dollars" & race != "All") %>%
  select(year, race, income_quintile, income_dollars)

# RIDGELINE PLOT
income_curr %>%
  ggplot(aes(x = income_dollars, y = income_quintile, fill = race)) +
    geom_density_ridges() +
    facet_grid(race ~ .) +
    scale_fill_manual(values = c("#5D295F", "#003685", "#78BE21", "#008EAA", "#FFC845")) +
    theme_mmb_basic(legend.position = "none",
                    plot.title.position = "plot",
                    axis.title.y = element_text(size = 12),
                    panel.grid.major.y = element_line(),
                    plot.caption = element_text(face = "italic", size = 12)
                    ) +
    labs(title = "Distribution of income by race (current dollars)",
         subtitle = "Years 1967 - 2019",
         caption = "Source: https://apps.urban.org/features/wealth-inequality-charts",
         x = "Income ($)",
         y = "")
```